---
title: "Veteran Disability and Economic Vulnerability: Assessing Poverty Risk through Federal Compensation and Individual Characteristics"
author: "Hau Phan, Kim By Goytia, Nancy Sandoval"
date: "2026-04-15"
output: pdf_document
bibliography: ref.bib
citeproc: true
csl: apa.csl
link-citations: true
editor_options: 
  markdown:
    wrap: none
header-includes:
  - \usepackage{indentfirst}
  - \setlength{\parindent}{2em}
  - \setlength{\parskip}{1em}
  - \usepackage{float}
  - \usepackage{placeins}
  
---

```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
library(readr)
library(psych)
library(dplyr)
library(tidyr)
library(broom)
library(knitr)
library(kableExtra)
library(ggplot2)
library(Hmisc)

pppub25 <- read_csv("pppub25.csv")
nrow(pppub25)

cleaned_pppub25 <- na.omit(pppub25)
nrow(cleaned_pppub25)
```

## 1. Introduction 

Amid increasing political and social concern about the complex challenges veterans face after military discharge, such as inadequate government support, mental health struggles, disability, and homelessness, our research seeks to examine the financial realities of disabled veterans in the United States. 

The motivation towards this research stems from prior studies that focused on the effect of denied veterans' affairs disability compensation on the health and social isolation of American Veterans. A study claims that American veterans who were denied veterans affairs disability compensation experienced deep poverty and isolation [@Fried2017HealthIsolation]. Those who were denied were most likely to be recipients of public assistance, unmarried, and have limited knowledge of VA benefits. This suggests they need more supportive services. Their research method consists of building 3 separate multivariate logistic regression models for three different domains of health to examine their unique effects. Another study on Risk Factors for Homelessness Among US Veterans finds that the strongest risk factors were substance abuse and mental illness, followed by extreme poverty and low-income-related factors(military pay and unemployment). Case studies found that there was a greater risk among veterans compared to the general population, and specifically populations from low-income backgrounds. The study also concerns the increasing number of homeless female veterans, for whom further research is needed on risk factors within these special subgroups [@NCHV_VeteranHomelessness]. The study supports policies and programs in efforts to end homelessness among U.S. veterans. According to the National Coalition for Homeless Veterans, additional factors influencing homelessness are due to the extreme shortage of affordable housing, livable income, and access to healthcare [@Tsai2015RiskHomelessness]. Homeless veterans also faced a lack of family and social support networks. Women veterans were also more likely to face homelessness after military separation. Additionally, 51% of homeless veterans in the demographics have disabilities. 

Our research questions **(1) Is veteran status associated with lower poverty level? Is disability status associated with lower poverty level?; (2) whether or not veteran status moderates disability status, racial, age, and gender disparities in poverty status; (3) among disabled people, whether current government support and disability compensation effectively protect them from poverty?** 
 
## 2. Background

Growing numbers of researchers find that Veteran’s affairs (VA) disability benefits play a significant role in shaping veterans’ economic stability and housing outcomes, yet important gaps remain regarding how these benefits translate into broader poverty risk. Across multiple studies, disability compensation functions as a stabilizing income floor that reduces material hardship. Evidence from mental disability compensation shows that higher benefit levels decrease homelessness, food insecurity, and delinquent debt, suggesting that direct cash transfers help secure basic needs [@silver2026invisible]. Similarly, service-connected disability ratings are associated with substantially lower homelessness risk among veterans receiving VA mental health services [@edens2011association]. However, this protective effect is not universal. Veterans discharged due to disability experience significantly elevated long-term homelessness risk, and benefits appear less protective for this subgroup compared to routinely discharged veterans [@FARGO2017592]. These mixed findings suggest that the economic effects of disability may depend in part on the level of veterans’ payments received. Therefore, we hypothesize **H1: Veteran payment and disability status does affect poverty status.**. 

An analysis of various papers centered around veteran status, disability, gender, and education level, as well as other compounding factors such as race and sexual orientation, reveals the emergence of common themes. Main themes and shared conclusions are: lack of understanding veteran identities post-service, especially intersectional identities [@blackstock2012gender], in studied employment cases, veteran identity serving as social capital regardless of disability status, and disability often functions as a greater barrier in academic and employment settings than gender alone [@ameri2025intersecting], and the large role disability plays in an individual's ability to self advocate[@kinney2017measuring]. Collectively, the papers reveal how, although veteran status can be concluded to be a social advantage, the intersectional identity of disability, sex, race, and age takes precedent in shaping access, belonging, and overall performance and success. 

For individuals with the intersecting identities: women of color veterans, sexual/gender minority veterans, and, in some cases, disability, feelings of invisibility, institutional neglect, and lack of belonging are intensified by these layered identities [@blackstock2012gender]. Overall, across higher education environments and employment contexts, disability serves as a large determinant of veteran success and outcome, with a particular emphasis on self-efficacy [@kinney2017measuring]. Many researchers have also examined income and health disparities among veterans from racial/ethnic minorities and younger or older groups. Study found that minority veterans received lower disability compensation approval rates than white veterans (@JPAG21483). Disability compensation was also prevalent among younger (ages 18–39) and older (ages 50–64) veterans (@BenShalom2016). As one might expect, military service provides individuals from disadvantaged backgrounds with opportunities to build their income. Aged Black and Hispanic veterans exhibit lower employment and higher work-limiting disability (@Tamborini2019). However, this study includes only male veterans and non-veterans. 

Another study on Iraq/Afghanistan-era veterans found that female and black veterans earn more than their non-veteran counterparts, which supports the advantages of veteran status from groups that generally have lower earnings (@Vick2017). However, Black adults, especially in the younger age group, had higher rates of homelessness among both veterans and nonveterans. Although these patterns are consistent in the general population and the poverty population, veteran status was linked to a greater risk of homelessness(@Fargo2012). In addition, minority veterans received less income and education than White veterans (@Sheehan2015). Furthermore, a study found that Black veterans received less favorable ratings on disability examinations than white veterans. The racial disparity in PTSD disability can directly cause a higher burden of poverty among Black veterans compared to other veterans (@Rosen2013). Furthermore, evidence from Veteran Health Administration(VHA)'s equal-access system does reduce disparities, but does not erase them entirely. There was evidence of mortality disparities in non-Hispanic Black men relative to non-Hispanic White men in both the VHA and U.S. populations. On the contrary, VHA non-Hispanic Black women did not experience mortality disparity relative to non-Hispanic Black women in the general U.S. population(@Wong2019). Due to not all veterans receiving veteran payments due to inequity, we will focus on veteran status in general instead of veteran payment to evalulate how veteran status interaction with demographic variables affect poverty. 

Lastly, another study found that Black and Hispanic veterans were more likely to have traumatic brain injury from assault/gun, but White veterans had higher VHA resource costs associated with assault/gun injuries(@DismukeGreer2021). 

Overall, these studies suggest that while veteran status may provide certain social and economic advantages, disparities across race, age, and gender persist among veterans. These findings indicate that the economic benefits associated with military service may vary across demographic groups. Therefore, we hypothesize **H2: veteran status moderates racial, disability status, age, and gender disparities in poverty status**.

Additional scholarship highlights how policy expansions, structural barriers, and eligibility rules shape both economic and behavioral outcomes for people with disabilities. People with disabilities have long experienced disproportionate levels of poverty and labor market exclusion; early research describes disabled individuals as “the poorest, least educated, and largest minority population in America” and emphasizes that employment barriers play a central role in persistent economic disadvantage [@Kopels01091995]. Broader sociological work similarly argues that disability is frequently linked to poverty through systemic exclusion from stable employment and labor markets, rather than solely through individual health limitations [@balazadeh2003disability]. Contemporary empirical research reinforces this relationship. For example, analyses using national survey data find that disabled adults are significantly more likely to live in poverty than their nondisabled counterparts, reflecting both reduced employment opportunities and institutional barriers within social programs [@heyman2024demographic]. In particular, parents with disabilities experience especially high poverty risk: roughly 27.3% live below the federal poverty line, compared with 11.4% of parents without disabilities [@heyman2024demographic], and these disparities persist even after adjusting for sociodemographic characteristics and participation in public assistance programs.

Policy expansions and eligibility rules also shape both economic and behavioral outcomes. Expanding compensability for specific service-related conditions increased diagnosis and health investment, demonstrating that financial incentives influence veterans’ engagement with the benefits system [@singleton2009effect]. At the same time, increased disability compensation has been shown to reduce traditional labor force participation while encouraging shifts into self-employment, reflecting income effects rather than moral hazard [@coile2021work]. Despite the rapid growth of the VA disability compensation program and its emergence as one of the largest federal disability systems, scholars emphasize a substantial lack of causal research evaluating its broader economic consequences [@armour2025veteran]. Administrative complexities further complicate access to benefits, as homeless veterans may receive expedited decisions yet remain less likely to be awarded service connection [@beydoun2025role], and differences between VA and Social Security disability standards create additional inconsistencies in income protection [@muller2014veterans].

Taken together, the literature suggests that disability status is closely linked to elevated poverty risk due to structural labor market barriers and institutional complexities within benefit systems. While disability compensation and public assistance programs may provide partial income protection, it remains unclear whether these supports substantially reduce poverty risk among disabled individuals overall. Therefore, we hypothesize **H3: Among disabled people, current government support and disability compensation help protect against poverty**.

```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
#Education
cleaned_pppub25 <- cleaned_pppub25 |>
  mutate(
    A_HGA_7cat = case_when(
      A_HGA %in% 0:34 ~ "Less than High School",
      A_HGA %in% 35:38 ~ "High School",
      A_HGA == 39 ~ "High School Graduate",
      A_HGA == 40 ~ "Some College",
      A_HGA %in% 41:42 ~ "Associate's Degree",
      A_HGA == 43 ~ "Bachelor's Degree",
      A_HGA %in% 44:46 ~ "Advanced Degree",
      TRUE ~ NA_character_
    ),
    A_HGA_7cat = factor(
      A_HGA_7cat,
      levels = c(
        "Less than High School",
        "High School",
        "High School Graduate",
        "Some College",
        "Associate's Degree",
        "Bachelor's Degree",
        "Advanced Degree"
      )
    )
  )
```


```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
#Region
cleaned_pppub25 <- cleaned_pppub25 |>
  mutate(
    MIG_REG = factor(
      MIG_REG,
      levels = c(1, 2, 3, 4, 5),
      labels = c("Northeast", "Midwest", "South", "West", "Abroad")
    )
  )

table(cleaned_pppub25$MIG_REG, useNA = "ifany")
```


```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}

#Employment
cleaned_pppub25 <- cleaned_pppub25 |>
  mutate(
    A_LFSR_3cat = case_when(
      A_LFSR %in% c("1", "2") ~ "Employed",
      A_LFSR %in% c("3", "4") ~ "Unemployed",
      A_LFSR == "7" ~ "Not in labor force",
      TRUE ~ NA_character_
    ),
    A_LFSR_3cat = factor(
      A_LFSR_3cat,
      levels = c("Employed", "Unemployed", "Not in labor force")
    )
  )
```

```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
#Marital
cleaned_pppub25 <- cleaned_pppub25 |>
  mutate(
    A_MARITL_3cat = case_when(
      A_MARITL %in% c("1", "2") ~ "Married",
      A_MARITL %in% c("3", "4", "5", "6") ~ "Previously married",
      A_MARITL == "7" ~ "Never married",
      TRUE ~ NA_character_
    ),
    A_MARITL_3cat = factor(
      A_MARITL_3cat,
      levels = c("Married", "Previously married", "Never married")
    )
  )
```


```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
#recoding VET_VAL into binary (0,1)
cleaned_pppub25$binVET_VAL <- factor(
  ifelse(cleaned_pppub25$VET_VAL > 0, 1, 0),
  levels = c(0, 1)
)

unique(cleaned_pppub25$binVET_VAL)
```


```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
# converting A_SEX variable into binary 
cleaned_pppub25$eA_SEX <- ifelse(cleaned_pppub25$A_SEX == 2, 1, 0)
unique(cleaned_pppub25$eA_SEX)
```

```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
#converting PEAFEVER variable into binary variable
cleaned_pppub25$ePEAFEVER <- ifelse(
  cleaned_pppub25$PEAFEVER == 1, 1,
  ifelse(cleaned_pppub25$PEAFEVER == 2, 0, NA)
)

table(cleaned_pppub25$ePEAFEVER, useNA = "ifany")
```


```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
# factor the numeric variables
cleaned_pppub25 <- cleaned_pppub25 |>
  mutate(
    A_HGA = factor(A_HGA),
    A_MARITL = factor(A_MARITL),
    A_LFSR = factor(A_LFSR),
    MIG_REG = factor(MIG_REG),
    PRDTRACE = factor(PRDTRACE), 
    eA_SEX = factor(eA_SEX)
  )

sapply(cleaned_pppub25[c("A_HGA", "A_MARITL", "A_LFSR", "MIG_REG", "PRDTRACE", "eA_SEX" )], class)
```

```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
# clean outcome and disability
cleaned_pppub25 <- cleaned_pppub25 |>
  filter(PERLIS != -1, PRDISFLG %in% c(1, 2), !is.na(ePEAFEVER)) |>
  mutate(
    PERLIS = if_else(PERLIS == 1, 1, 0),
    ePRDISFLG = ifelse(PRDISFLG == 1, 1, 0)
  )

unique(cleaned_pppub25$PERLIS)
class(cleaned_pppub25$PERLIS)
table(cleaned_pppub25$ePRDISFLG, useNA = "ifany")
```


```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
cleaned_pppub25 <- cleaned_pppub25 |>
  mutate(
    PRDTRACE_num = as.numeric(as.character(PRDTRACE)),
    race_ethnicity = case_when(
      PEHSPNON == 1 ~ "Hispanic only",
      PEHSPNON == 2 & PRDTRACE_num == 1 ~ "Non-Hispanic White only",
      PEHSPNON == 2 & PRDTRACE_num == 2 ~ "Non-Hispanic Black only",
      PEHSPNON == 2 & PRDTRACE_num %in% c(4, 5) ~ "Non-Hispanic Asian only",
      PEHSPNON == 2 & PRDTRACE_num == 3 ~ "Non-Hispanic Indigenous only",
      PEHSPNON == 2 & PRDTRACE_num %in% 6:26 ~ "Non-Hispanic Multiracial/Other",
      TRUE ~ NA_character_
    ),
    race_ethnicity = factor(
      race_ethnicity,
      levels = c(
        "Non-Hispanic White only",
        "Non-Hispanic Black only",
        "Non-Hispanic Asian only",
        "Non-Hispanic Indigenous only",
        "Non-Hispanic Multiracial/Other",
        "Hispanic only"
      )
    )
  )

table(cleaned_pppub25$race_ethnicity, useNA = "ifany")
table(cleaned_pppub25$PEHSPNON, useNA = "ifany")
```

```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
#Model 1
ProjectModel1 <- cleaned_pppub25 |>
  dplyr::select(
    PERLIS,
    binVET_VAL,
    ePRDISFLG,
    A_AGE,
    eA_SEX,
    A_MARITL_3cat,
    A_HGA_7cat,
    race_ethnicity,
    MIG_REG,
    A_LFSR_3cat
  ) |>
  na.omit()

ProjectModel1 <- ProjectModel1 |>
  mutate(
    binVET_VAL = factor(
      binVET_VAL,
      levels = c(0, 1),
      labels = c("No veteran payment", "Veteran payment")
    ),
    ePRDISFLG = factor(
      ePRDISFLG,
      levels = c(0, 1),
      labels = c("No disability", "Disability")
    ),
    eA_SEX = factor(
      eA_SEX,
      levels = c(0, 1),
      labels = c("Female", "Male")
    ),
    A_MARITL_3cat = relevel(A_MARITL_3cat, ref = "Married"),
    A_HGA_7cat = relevel(A_HGA_7cat, ref = "Less than High School"),
    race_ethnicity = relevel(race_ethnicity, ref = "Non-Hispanic White only"),
    A_LFSR_3cat = relevel(A_LFSR_3cat, ref = "Employed")
  )
```

## 3. Methods
### 3.1 Data Source 
  The data we use for this research come from the CPS-Annual Social and Economic Supplement (ASEC) data [@Census_CPSASEC] for 2025 (with additional data years) from the Current Population Survey (CPS) [@Census_CPS], which is the source of official government statistics on employment and unemployment. For more than 50 years, CPS has been conducted every month. Approximately 54,000 households are interviewed each month to represent the country, states, and other specific locations. Each home is interviewed once a month for four months in a row, leaves the sample, and is then interviewed again for the same duration the following year. The data provide detailed work experience on job status, industry, and occupation for individuals age 15 and older. The CPS includes household, family, and person-level records. In this study, we focus on the person-level record to retrieve specific information such as disability status, veteran status, poverty status, and demographic characteristics.
  
  To handle missing data, we excluded observations with missing values on variables included in each model. For Models 1 and 3, we retained both disabled and non-disabled respondents, resulting in a sample size of 109,707 respondents. For Model 2, we restricted the sample to disabled respondents only, resulting in a sample size of 14,204 respondents. In addition, we excluded individuals who responded as being not eligible to be evaluated for poverty status.


The explanatory variables we use include:

1\. A_AGE: Age, treated as a continuous variable.

2\. A_HGA: Educational level, grouped into seven categories: Less than High School, High School, High School Graduate, Some College, Associate’s Degree, Bachelor’s Degree, and Advanced Degree. This variable is treated as categorical, with Less than High School as the reference group.

3\. A_MARITL: Marital status, treated as a categorical variable with Married as the reference group.

4\. A_SEX: Sex, originally coded numerically and recoded into a binary variable where 0 = female and 1 = male, with female as the reference group.

5\. A_LFSR: Employment status, treated as a categorical variable with Employed as the reference group.

6\. DSAB_VAL: Disability income, treated as a continuous variable.

7\. MIG_REG: Region, treated as a categorical variable with Northeast as the reference group.

8\. PEAFEVER: Veteran status, recoded into a binary variable where 0 = not a veteran and 1 = veteran.

9\. race_ethnicity: Combined race/ethnicity variable created from PRDTRACE and PEHSPNON, including Hispanic only, Non-Hispanic White only, Non-Hispanic Black only, Non-Hispanic Asian only, Non-Hispanic Indigenous only, and Non-Hispanic Multiracial/Other. The reference group is Non-Hispanic White only.

10\. PRDISFLG: Disability status (binary variable coded 0 = no disability and 1 = disability; reference group = no disability).

11\. VET_VAL: Veteran payment status, originally recorded as veteran-related income amount and recoded into a binary variable where 0 = no veteran payment received and 1 = veteran payment received, with no veteran payment as the reference group.

And our outcome variable for all three models is:

1\. PERLIS: CPS person-level poverty indicator, recoded into a binary outcome where 1 = in poverty and 0 = not in poverty. Observations with PERLIS = -1 were excluded because they were not eligible to be evaluated for poverty.

From these data, we estimate logistic regression models predicting poverty status using CPS ASEC person-level records. The outcome variable is based on the 2025 federal poverty threshold and is measured using the CPS person-level poverty indicator (PERLIS), coded 1 if an individual falls below the federal poverty line and 0 otherwise.

### 3.2 Stepwise Regression Test

```{r, results='asis', echo = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
model1_additive <- glm(
  PERLIS ~ binVET_VAL + ePRDISFLG + A_AGE + eA_SEX +
    A_MARITL_3cat + A_HGA_7cat + race_ethnicity + MIG_REG + A_LFSR_3cat,
  data = ProjectModel1,
  family = binomial()
)

model1_interaction <- glm(
  PERLIS ~ binVET_VAL * ePRDISFLG + A_AGE + eA_SEX +
    A_MARITL_3cat + A_HGA_7cat + race_ethnicity + MIG_REG + A_LFSR_3cat,
  data = ProjectModel1,
  family = binomial()
)

lrt_model1 <- anova(model1_additive, model1_interaction, test = "Chisq")

model1_testing_table <- data.frame(
  Models = c(
    "Model 1A: VET_VAL + Disability + covariates",
    "Model 1B: VET_VAL * Disability + covariates"
  ),
  ANOVA = c(
    "-",
    ifelse(
      is.na(lrt_model1$`Pr(>Chi)`[2]),
      "-",
      formatC(lrt_model1$`Pr(>Chi)`[2], format = "f", digits = 4)
    )
  ),
  AIC = c(
    round(AIC(model1_additive), 1),
    round(AIC(model1_interaction), 1)
  )
)

knitr::kable(
  model1_testing_table,
  format = "latex",
  booktabs = TRUE,
  caption = "Model Comparison for Model 1",
  col.names = c("Models", "ANOVA", "AIC")
) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position"))
```

Table 1 compares the additive and interaction specifications for Model 1. The additive model had a slightly lower AIC (58182.5) than the interaction model (58184.2). In addition, the likelihood ratio test comparing the two models was not statistically significant, p = 0.5481. Together, these results suggested that adding the interaction between veteran payment status and disability status did not meaningfully improve model fit, so we retained the additive specification for Model 1.

Model 1 estimates the main effects associated with poverty in the general population. The primary explanatory variables in this model are veteran payment status, measured using the binary veteran payment variable derived from VET_VAL, and disability status, measured using PRDISFLG. The model also includes age, sex, and race/ethnicity as additional explanatory variables of interest. Education, marital status, region, and employment status are included as covariates. Therefore, Model 1 assesses whether veteran payment status and disability status are independently associated with poverty in the general population while accounting for key demographic and socioeconomic characteristics.

The first model is estimated as: 

$$
\begin{aligned}
\text{logit}\big(P(PERLIS_i = 1)\big) =\;& \beta_0 \\
&+ \beta_1 PRDISFLG_i \\
&+ \beta_2 VET\_VAL_i \\
&+ \beta_3 A\_AGE_i \\
&+ \beta_4 race\_ethnicity_i \\
&+ \beta_5 A\_SEX_i \\
&+ \beta_6 A\_MARITL_i \\
&+ \beta_7 MIG\_REG_i \\
&+ \beta_8 A\_LFSR_i \\
&+ \beta_{9} A\_HGA_i \\
&+ \varepsilon_i
\end{aligned}
$$

```{r, include = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
#model 3 

ProjectModel3 <- cleaned_pppub25 |>
  dplyr::select(PERLIS, eA_SEX, A_AGE, race_ethnicity, ePEAFEVER, A_HGA_7cat, A_MARITL_3cat, A_LFSR_3cat, MIG_REG, ePRDISFLG) |>
  na.omit()

psych::describe(ProjectModel3)
```

```{r, include = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
# base model

base_model <- glm(
 PERLIS ~ A_HGA_7cat + A_MARITL_3cat + A_LFSR_3cat + MIG_REG,
 data = ProjectModel3,
 family = binomial()
)

# fit model for sex/veteran status interaction
model_sex <- glm(
  PERLIS ~ eA_SEX + ePEAFEVER +
    eA_SEX:ePEAFEVER +
    A_HGA_7cat + A_MARITL_3cat + A_LFSR_3cat + MIG_REG,
  data = ProjectModel3,
  family = binomial()
)

# fit model for age/veteran status interaction
model_age <- glm(
 PERLIS ~ A_AGE + ePEAFEVER + A_AGE:ePEAFEVER +
   A_HGA_7cat + A_MARITL_3cat + A_LFSR_3cat + MIG_REG,
 data = ProjectModel3,
 family = binomial()
)

# fit model for race/veteran status interaction
model_race <- glm(
  PERLIS ~ race_ethnicity + ePEAFEVER +
    race_ethnicity:ePEAFEVER +
    A_HGA_7cat + A_MARITL_3cat + A_LFSR_3cat + MIG_REG,
  data = ProjectModel3,
  family = binomial()
)

# fit model for disability/veteran status interaction
model_dis <- glm(
  PERLIS ~ ePRDISFLG + ePEAFEVER +
    ePRDISFLG:ePEAFEVER +
    A_HGA_7cat + A_MARITL_3cat + A_LFSR_3cat + MIG_REG,
  data = ProjectModel3,
  family = binomial()
)

```

```{r, include = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}

# do anova and AIC test
anova(base_model, model_sex, test = "Chisq")
anova(base_model, model_age, test = "Chisq") # significant 
anova(base_model, model_race, test = "Chisq")
anova(base_model, model_dis, test = "Chisq")

AIC(base_model, model_sex, model_age, model_race, model_dis)
```

```{r, include = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}

# model with Race/veteran status interaction and Age/veteran status interaction
model_2 <- glm(
 PERLIS ~ race_ethnicity + ePEAFEVER + race_ethnicity:ePEAFEVER + A_AGE + A_AGE:ePEAFEVER +
   A_HGA_7cat + A_MARITL_3cat + A_LFSR_3cat + MIG_REG,
 data = ProjectModel3,
 family = binomial()
)

# model with Race/veteran status interaction, Age/veteran status interaction, and Sex/veteran status interaction
model_3 <- glm(
 PERLIS ~ race_ethnicity + ePEAFEVER + race_ethnicity:ePEAFEVER + A_AGE + A_AGE:ePEAFEVER + eA_SEX + eA_SEX:ePEAFEVER + A_HGA_7cat + A_MARITL_3cat + A_LFSR_3cat + MIG_REG,
 data = ProjectModel3,
 family = binomial()
)

# model with Race/veteran status interaction, Age/veteran status interaction, Sex/veteran status interaction, and disability status/veteran status interaction
model_4 <- glm(
 PERLIS ~ race_ethnicity + ePEAFEVER + race_ethnicity:ePEAFEVER + A_AGE + A_AGE:ePEAFEVER + eA_SEX + eA_SEX:ePEAFEVER + ePRDISFLG + ePRDISFLG:ePEAFEVER + 
   A_HGA_7cat + A_MARITL_3cat + A_LFSR_3cat + MIG_REG,
 data = ProjectModel3,
 family = binomial()
)

```

```{r, echo = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
model_testing_table <- data.frame(
  Models = c(
    "Model 1: Main effects only",
    "Model 2A: Race/Ethnicity × Veteran Status + covariates",
    "Model 2B: Race/Ethnicity × Veteran Status + Age × Veteran Status + covariates",
    "Model 2C: Race/Ethnicity × Veteran Status + Age × Veteran Status + Sex × Veteran Status + covariates",
    "Model 2D: Race/Ethnicity × Veteran Status + Age × Veteran Status + Sex × Veteran Status + Disability × Veteran Status + covariates"
  ),
  
  ANOVA = c(
    "-",
    
    ifelse(is.na(anova(base_model, model_age, test = "Chisq")$`Pr(>Chi)`[2]),
           "-",
           formatC(anova(base_model, model_age, test = "Chisq")$`Pr(>Chi)`[2],
                   format = "f", digits = 4)),
    
    ifelse(is.na(anova(model_age, model_2, test = "Chisq")$`Pr(>Chi)`[2]),
           "-",
           formatC(anova(model_age, model_2, test = "Chisq")$`Pr(>Chi)`[2],
                   format = "f", digits = 4)),
    
    ifelse(is.na(anova(model_2, model_3, test = "Chisq")$`Pr(>Chi)`[2]),
           "-",
           formatC(anova(model_2, model_3, test = "Chisq")$`Pr(>Chi)`[2],
                   format = "f", digits = 4)),
    
    ifelse(is.na(anova(model_3, model_4, test = "Chisq")$`Pr(>Chi)`[2]),
           "-",
           formatC(anova(model_3, model_4, test = "Chisq")$`Pr(>Chi)`[2],
                   format = "f", digits = 4))
  ),
  
  AIC = c(
    round(AIC(base_model), 1),
    round(AIC(model_age), 1),
    round(AIC(model_2), 1),
    round(AIC(model_3), 1),
    round(AIC(model_4), 1)
  )
)

knitr::kable(
  model_testing_table,
  format = "latex",
  booktabs = TRUE,
  caption = "Model Comparison for Poverty Models",
  col.names = c("Models", "ANOVA", "AIC"),
  escape = FALSE
) |>
  kableExtra::kable_styling(
    latex_options = c("HOLD_position"),
    full_width = FALSE
  ) |>
  kableExtra::column_spec(1, width = "10cm")
```
Table 2 compares the main-effects model with models that add veteran status interactions related to our research questions. Model 2A, which includes the Race/Ethnicity × Veteran Status interaction, had the lowest AIC among the models tested. Model 2B added the Age × Veteran Status interaction, Model 2C added the Sex × Veteran Status interaction, and Model 2D added the Disability Status × Veteran Status interaction. Because Model 2D had the lowest AIC, we selected the model with all of the interactions as the finalized Model 2.
				
The finalized second model assesses whether veteran status moderates disability status, racial, age, and gender disparities in poverty status. The primary explanatory variables in the finalized model are veteran status, age, sex, race/ethnicity, and disability status including the interaction between veteran status and age, the interaction between veteran status and sex, the interaction between veteran status and race, and the interaction between veteran status and disability status. Education, marital status, employment status, and region are included as covariates. 

The second model is estimated as: 

$$
\begin{aligned}
\text{logit}\big(P(PERLIS_i = 1)\big) =\;& \beta_0 \\
&+ \beta_1 race\_ethnicity_i \\
&+ \beta_2 ePEAFEVER_i \\
&+ \beta_3 (race\_ethnicity_i \times ePEAFEVER_i) \\
&+ \beta_4 A\_AGE_i \\
&+ \beta_5 (A\_AGE_i \times ePEAFEVER_i) \\
&+ \beta_6 eA\_SEX_i \\
&+ \beta_7 (eA\_SEX_i \times ePEAFEVER_i) \\
&+ \beta_8 ePRDISFLG_i \\
&+ \beta_9 (ePRDISFLG_i \times ePEAFEVER_i) \\
&+ \beta_{10} A\_HGA\_7cat_i \\
&+ \beta_{11} A\_MARITL\_3cat_i \\
&+ \beta_{12} A\_LFSR\_3cat_i \\
&+ \beta_{13} MIG\_REG_i
\end{aligned}
$$

The third model focuses on the financial status of disabled individuals, including both disabled veterans and disabled civilians. For this model, the analytic sample is restricted to disabled respondents by using PRDISFLG = 1 for disability status. In addition to demographic controls, we include person-level measures of veterans’ payments (VET_VAL) and disability income (DSAB_VAL) to assess whether current government support and disability compensation protect against poverty among disabled people. We also include veteran status, measured using PEAFEVER.

The third model is estimated as:

$$
\begin{aligned}
\text{logit}\big(P(PERLIS_i = 1)\big) =\;& \beta_0 \\
&+ \beta_1 A\_SEX_i \\
&+ \beta_2 VET\_VAL_i \\
&+ \beta_3 DSAB\_VAL_i \\
&+ \beta_4 A\_AGE_i \\
&+ \beta_5 race\_ethnicity_i \\
&+ \beta_6 A\_HGA_i \\
&+ \beta_7 A\_MARITL_i \\
&+ \beta_8 A\_LFSR_i \\
&+ \beta_9 MIG\_REG_i \\
&+ \beta_{10} PEAFEVER_i \\
&+ \varepsilon_i
\end{aligned}
$$

All three models focus on person-level variables for ease of interpretation. By examining these relationships, this study aims to determine whether current government support and disability compensation effectively protect disabled veterans from poverty or whether economic gaps persist following military discharge.

```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
ProjectModel1_table <- ProjectModel1 |>
  mutate(
    PERLIS_label = factor(
      PERLIS,
      levels = c(0, 1),
      labels = c("Not in poverty", "In poverty")
    )
  )
```


```{r include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
model1_add <- glm(
  PERLIS ~ binVET_VAL + ePRDISFLG + A_AGE + eA_SEX +
    A_MARITL_3cat + A_HGA_7cat + race_ethnicity + MIG_REG + A_LFSR_3cat,
  data = ProjectModel1,
  family = binomial()
)

model1_simple <- glm(
  PERLIS ~ binVET_VAL + ePRDISFLG,
  data = ProjectModel1,
  family = binomial()
)

model1_interactive <- glm(
  PERLIS ~ binVET_VAL + ePRDISFLG + binVET_VAL*ePRDISFLG+ A_AGE + eA_SEX +
    A_MARITL_3cat + A_HGA_7cat + race_ethnicity + MIG_REG + A_LFSR_3cat,
  data = ProjectModel1,
  family = binomial()
)

summary(model1_add)
summary(model1_interactive)

anova(model1_add,model1_interactive)
```

## 4. Results 

### 4.1 Descriptive Statistics
```{r echo = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
ModelTesting_Data <- cleaned_pppub25 |>
  dplyr::select(
    PERLIS,
    binVET_VAL,
    ePRDISFLG,
    A_AGE,
    eA_SEX,
    DSAB_VAL,
    ePEAFEVER, 
  ) |>
  na.omit() |>
  mutate(
    PERLIS = as.numeric(PERLIS),
    binVET_VAL = as.numeric(as.character(binVET_VAL)),
    ePRDISFLG = as.numeric(ePRDISFLG),
    A_AGE = as.numeric(A_AGE),
    eA_SEX = as.numeric(eA_SEX),
    DSAB_VAL = as.numeric(DSAB_VAL),
    ePEAFEVER = as.numeric(ePEAFEVER)
  )

correlation_matrix <- rcorr(as.matrix(ModelTesting_Data), type = "pearson")

des_stats_correlations_table <- function(Correlation_Matrix, Data_Matrix, DVindex, Data){
  R <- Correlation_Matrix$r
  p <- Correlation_Matrix$P
  mystars <- ifelse(p < .001, "***", ifelse(p < .01, "**", ifelse(p < .05, "*", "")))
  R <- round(R, digits = 2)
  Rnew <- matrix(paste0(R, mystars), ncol = ncol(Data_Matrix))
  diag(Rnew) <- paste0(diag(R))
  rownames(Rnew) <- colnames(Data_Matrix)
  colnames(Rnew) <- colnames(Data_Matrix)
  Rnew_df <- as.data.frame(Rnew)

  row_to_move <- DVindex
  new_order <- c(row_to_move, setdiff(1:nrow(Rnew_df), row_to_move))
  Rnew_df <- Rnew_df[new_order, ]
  col_to_remove <- DVindex
  Rnew_df <- Rnew_df[, -col_to_remove]

  Rnew <- as.matrix(Rnew_df)
  Rnew_DV <- Rnew[1, ]
  Rnew_IV <- Rnew[-1, ]
  Rnew_IV[upper.tri(Rnew_IV, diag = TRUE)] <- "-"
  Rnew_IV <- as.data.frame(Rnew_IV)
  Rnew_df <- rbind(Rnew_DV, Rnew_IV)
  rownames(Rnew_df)[1] <- colnames(Data_Matrix)[DVindex]

  des_stats <- sapply(Data, function(x) c(M = mean(x, na.rm = TRUE), SD = sd(x, na.rm = TRUE)))
  des_stats <- t(as.data.frame(des_stats))
  des_stats <- round(des_stats, digits = 2)
  des_stats <- des_stats[new_order, ]

  cbind(des_stats, Rnew_df)
}

DVindex <- 1

testing_table1 <- des_stats_correlations_table(
  correlation_matrix,
  as.matrix(ModelTesting_Data),
  DVindex,
  ModelTesting_Data
)

rownames(testing_table1) <- c(
  "Poverty status",
  "1. Veteran payment",
  "2. Disability status",
  "3. Age",
  "4. Sex",
  "5. Disability Income",
  "6. Veteran status"
)

colnames(testing_table1) <- c("M", "SD", "1", "2", "3", "4", "5", "6")

kable(
  testing_table1,
  format = "latex",
  booktabs = TRUE,
  caption = "Means, Standard Deviations, and Intercorrelations for Model Testing Variables"
) |>
  kable_styling(latex_options = c("HOLD_position", "scale_down")) |>
  footnote(general = "*p < .05, **p < .01, ***p < .001.",
 threeparttable = TRUE)
```

Table 3 presents means, standard deviations, and intercorrelations for the outcome variable, the primary explanatory variables, and selected covariates used in model testing. Overall, the correlations were small to moderate in magnitude, suggesting that the variables were not highly correlated and could be included in the same regression models. The largest observed association was between poverty status and veteran status, but this relationship was not large enough to indicate problematic multicollinearity. Based on this screening, we retained the variables for further modeling.


### 4.2 Model 1: Logistic Regression Predicting Poverty Status Among General Population

This model estimated a general population logistic regression predicting the odds of being in poverty. The analytic sample included respondents with complete data on poverty status, disability status, veteran payment status, age, sex, education, marital status, race/ethnicity, employment status, and region, resulting in a final sample of 109,707 respondents. 

```{r = table_4, echo = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
table4 <- broom::tidy(model1_additive, conf.int = TRUE, exponentiate = FALSE) |>
  mutate(
    OR = exp(estimate),
    CI_lower = exp(conf.low),
    CI_upper = exp(conf.high),
    term = recode(
      term,
      "(Intercept)" = "Intercept",
      "binVET_VALVeteran payment" = "Veteran payment",
      "ePRDISFLGDisability" = "Disability",
      "A_AGE" = "Age",
      "eA_SEXMale" = "Male",
      "A_MARITL_3catPreviously married" = "Previously married",
      "A_MARITL_3catNever married" = "Never married",
      "A_HGA_7catHigh School" = "High School",
      "A_HGA_7catHigh School Graduate" = "High School Graduate",
      "A_HGA_7catSome College" = "Some College",
      "A_HGA_7catAssociate's Degree" = "Associate's Degree",
      "A_HGA_7catBachelor's Degree" = "Bachelor's Degree",
      "A_HGA_7catAdvanced Degree" = "Advanced Degree",
      "race_ethnicityNon-Hispanic Black only" = "Non-Hispanic Black only",
      "race_ethnicityNon-Hispanic Asian only" = "Non-Hispanic Asian only",
      "race_ethnicityNon-Hispanic Indigenous only" = "Non-Hispanic Indigenous only",
      "race_ethnicityNon-Hispanic Multiracial/Other" = "Non-Hispanic Multiracial/Other",
      "race_ethnicityHispanic only" = "Hispanic only",
      "MIG_REGMidwest" = "Region: Midwest",
      "MIG_REGSouth" = "Region: South",
      "MIG_REGWest" = "Region: West",
      "MIG_REGAbroad" = "Region: Abroad",
      "A_LFSR_3catUnemployed" = "Unemployed",
      "A_LFSR_3catNot in labor force" = "Not in labor force"
    ),
    B = round(estimate, 3),
    SE = round(std.error, 3),
    z = round(statistic, 3),
    p = ifelse(p.value < .001, "< .001", sprintf("%.3f", p.value)),
    OR = round(OR, 3),
    `95% CI for OR` = paste0("[", round(CI_lower, 3), ", ", round(CI_upper, 3), "]")
  ) |>
  select(term, B, SE, z, p, OR, `95% CI for OR`) |>
  rename(Predictor = term)

kable(
  table4,
  format = "latex",
  booktabs = TRUE,
  linesep = "",
  caption = "Logistic Regression Predicting Poverty Status in Model 1"
) |>
  kable_styling(latex_options = c("HOLD_position", "scale_down")) |>
  footnote(general = "OR = odds ratio.")
```

 Table 4 presents the logistic regression results for Model 1. Veteran payment status was negatively associated with poverty. Respondents receiving veteran-related payments had significantly lower odds of being in poverty than respondents not receiving veteran payments (OR = 0.236, 95% CI = [0.182,0.302], p < .001) while controlling for all other covariates. This indicated that veteran payment was associated with substantially reduced poverty odds in the general population model. Disability status was positively associated with poverty. Respondents with disabilities had significantly higher odds of poverty than respondents without disabilities (OR = 1.273, 95% CI = [1.202,1.348], p < .001).

 Several demographic and socioeconomic covariates also showed strong associations with poverty. Age was negatively associated with poverty (OR = 0.992, 95% CI [0.990, 0.993], p < .001), indicating that older respondents had slightly lower odds of poverty. Men had higher odds of poverty than women (OR = 1.181, 95% CI [1.130, 1.235], p < .001). Marital status was strongly associated with poverty. Compared with married respondents, those who were previously married had about 3.08 times the odds of poverty (OR = 3.080, 95% CI [2.904, 3.268], p < .001) and those who had never married had about 2.75 times the odds (OR = 2.745, 95% CI [2.577, 2.923], p < .001). 
	
Education also appeared protective against poverty. Compared with respondents with less than a high school education, all higher educational attainment groups were associated with significantly lower odds of poverty. Respondents with bachelor’s degrees had 71.9% lower odds of poverty (OR = 0.281, 95% CI [0.252, 0.313], p < .001), and respondents with an advanced degree had 78.1% lower odds of poverty (OR = 0.219, 95% CI [0.192, 0.249], p < .001).
	Race/ethnicity was also associated with poverty. Compared with Non-Hispanic White respondents, Non-Hispanic Black respondents had higher odds of poverty (OR = 1.688, 95% CI [1.585, 1.797], p < .001), as did Non-Hispanic Indigenous respondents (OR = 2.088, 95% CI [1.776, 2.446], p < .001), Hispanic respondents (OR = 1.484, 95% CI [1.403, 1.570], p < .001), and Non-Hispanic Asian respondents (OR = 1.127, 95% CI [1.025, 1.238], p = .013). The coefficient for Non-Hispanic Multiracial/Other respondents, however, was not statistically significant (OR = 1.081, 95% CI [0.923, 1.260], p = .329).
	
Employment was another strong predictor of poverty. Compared with employed respondents, unemployed respondents had nearly 3.95 times the odds of poverty (OR = 3.946, 95% CI [3.549, 4.382], p < .001), and respondents not in the labor force had about 4.71 times the odds (OR = 4.712, 95% CI [4.476, 4.962], p < .001). In relation to region, respondents living in the South or abroad had significantly higher odds of poverty than respondents in the Northeast. Overall, Model 1 showed that poverty in the general population was associated with disability status, veteran payment status, education, marital status, race/ethnicity, and labor force status. Although veteran payment was associated with lower poverty odds, disability remained associated with elevated poverty risk.

### 4.3 Model 2: Logistic Regression Predicting Poverty from Veteran Status Interactions

```{r, include = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
#model 3 

ProjectModel3 <- cleaned_pppub25 |>
  dplyr::select(PERLIS, eA_SEX, A_AGE, race_ethnicity, ePEAFEVER, A_HGA_7cat, A_MARITL_3cat, A_LFSR_3cat, MIG_REG, ePRDISFLG) |>
  na.omit()

psych::describe(ProjectModel3)
```
The second model estimates a general population logistic regression predicting the odds of being in poverty using veteran status interactions with age, sex, disability status, and race/ethnicity, while controlling for education, marital status, employment status, and region. The sample includes respondents with complete data on poverty status, veteran status, disability status, age, sex, education, marital status, race/ethnicity, employment status, and region.

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
library(ggeffects)
library(broom)
library(dplyr)
library(knitr)
library(kableExtra)

table3 <- broom::tidy(model_4, conf.int = TRUE, exponentiate = FALSE) |>
  mutate(
    OR = exp(estimate),
    CI_lower = exp(conf.low),
    CI_upper = exp(conf.high),
    term = recode(
      term,
      "(Intercept)" = "Intercept",
      
      # Main effects
      "eA_SEX1" = "Male",
      "A_AGE" = "Age",
      "ePEAFEVER" = "Veteran",
      "ePRDISFLG1" = "Disability",
      "ePRDISFLG" = "Disability",
      
      # Race/ethnicity
      "race_ethnicityNon-Hispanic Black only" = "Non-Hispanic Black only",
      "race_ethnicityNon-Hispanic Asian only" = "Non-Hispanic Asian only",
      "race_ethnicityNon-Hispanic Indigenous only" = "Non-Hispanic Indigenous only",
      "race_ethnicityNon-Hispanic Multiracial/Other" = "Non-Hispanic Multiracial/Other",
      "race_ethnicityHispanic only" = "Hispanic only",
      
      # Education
      "A_HGA_7catHigh School" = "High School",
      "A_HGA_7catHigh School Graduate" = "High School Graduate",
      "A_HGA_7catSome College" = "Some College",
      "A_HGA_7catAssociate’s Degree" = "Associate's Degree",
      "A_HGA_7catAssociate's Degree" = "Associate's Degree",
      "A_HGA_7catBachelor’s Degree" = "Bachelor's Degree",
      "A_HGA_7catBachelor's Degree" = "Bachelor's Degree",
      "A_HGA_7catAdvanced Degree" = "Advanced Degree",
      
      # Marital status
      "A_MARITL_3catPreviously married" = "Previously married",
      "A_MARITL_3catNever married" = "Never married",
      
      # Employment
      "A_LFSR_3catUnemployed" = "Unemployed",
      "A_LFSR_3catNot in labor force" = "Not in labor force",
      
      # Region
      "MIG_REGMidwest" = "Midwest",
      "MIG_REGSouth" = "South",
      "MIG_REGWest" = "West",
      "MIG_REGAbroad" = "Abroad",
      
      # Interaction effects
     "race_ethnicityNon-Hispanic Black only:ePEAFEVER" =
  "Non-Hispanic Black only × Veteran",
"race_ethnicityNon-Hispanic Asian only:ePEAFEVER" =
  "Non-Hispanic Asian only × Veteran",
"race_ethnicityNon-Hispanic Indigenous only:ePEAFEVER" =
  "Non-Hispanic Indigenous only × Veteran",
"race_ethnicityNon-Hispanic Multiracial/Other:ePEAFEVER" =
  "Non-Hispanic Multiracial/Other × Veteran",
"race_ethnicityHispanic only:ePEAFEVER" =
  "Hispanic only × Veteran",

"ePEAFEVER:A_AGE" = "Age × Veteran",
"ePEAFEVER:eA_SEX1" = "Male × Veteran",
"ePEAFEVER:ePRDISFLG" = "Disability × Veteran"
    ),
    B = round(estimate, 3),
    SE = round(std.error, 3),
    z = round(statistic, 3),
    p = ifelse(p.value < .001, "< .001", sprintf("%.3f", p.value)),
    OR = round(OR, 3),
    `95% CI for OR` = paste0("[", round(CI_lower, 3), ", ", round(CI_upper, 3), "]")
  ) |>
  select(term, B, SE, z, p, OR, `95% CI for OR`) |>
  rename(Predictor = term)

kable(
  table3,
  format = "latex",
  booktabs = TRUE,
  linesep = "",
  caption = "Logistic Regression Predicting Poverty Status in Model 2"
) |>
  kable_styling(latex_options = c("HOLD_position", "scale_down")) |>
  footnote(general = "OR = odds ratio")

```
When assessing the interaction terms, the Age × Veteran interaction was statistically significant, OR = 0.992, 95% CI [0.985, 0.998], p = .008. This suggests that the association between age and poverty differs by veteran status. Specifically, the odds of poverty associated with age were slightly lower among veterans than non-veterans, controlling for the other variables in the model. Therefore, there is evidence that veteran status moderates the relationship between age and poverty status.

Among the race/ethnicity interaction terms, the Hispanic only × Veteran interaction was also statistically significant, OR = 0.670, 95% CI [0.470, 0.937], p = .023. This suggests that the association between Hispanic ethnicity and poverty differs by veteran status. The other race/ethnicity interactions were not statistically significant at the .05 level, including Non-Hispanic Black only × Veteran, Non-Hispanic Asian only × Veteran, Non-Hispanic Indigenous only × Veteran, and Non-Hispanic Multiracial/Other × Veteran. The Male × Veteran and Disability × Veteran interactions were also not statistically significant.

Several main effects were also meaningful predictors of poverty status. Respondents with disabilities had about 27% higher odds of poverty than respondents without disabilities, OR = 1.269. Male respondents had about 18% higher odds of poverty than female respondents, OR = 1.178. Higher education was associated with lower odds of poverty: respondents with bachelor’s degrees had about 72% lower odds of poverty, OR = 0.281, and respondents with advanced degrees had about 78% lower odds of poverty, OR = 0.219, compared with respondents with less than a high school education. Marital status and labor force status were also strongly associated with poverty. Previously married respondents had about 3.1 times the odds of poverty, and never-married respondents had about 2.8 times the odds of poverty, compared with married respondents. Compared with employed respondents, unemployed respondents had about 3.9 times the odds of poverty, and respondents not in the labor force had about 4.7 times the odds of poverty.

Overall, this model suggests that veteran status moderates the relationship between age and poverty status, and may also moderate the relationship between Hispanic ethnicity and poverty status. However, most other veteran-status interactions were not statistically significant. Poverty risk was also strongly shaped by disability status, education, marital status, employment status, and race/ethnicity.

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
# graph for interaction between age and veteran status 
pred_age <- ggpredict(model_4, terms = c("A_AGE", "ePEAFEVER"))

ggplot(pred_age, aes(x = x, y = predicted, linetype = group)) +
  geom_line(linewidth = 1) +
  scale_linetype_discrete(
    labels = c("Non-veteran", "Veteran")
  ) +
  labs(
    x = "Age",
    y = "Predicted Probability of Poverty",
    linetype = "Veteran status",
    title = "Figure 1: Predicted Probability of Poverty by Age and Veteran Status"
  ) +
  coord_cartesian(ylim = c(0, 0.10)) +
  theme_minimal()

```

In Figure 1, the poverty level decreases with age for both veterans and non-veterans. This indicate that older respondents have lower odds of poverty than younger respondents. The line appears steeper for veterans than for non-veterans which indicates that veteran status does moderate the relationship between age and poverty status. The poverty level gap between veterans and non-veterans appear larger for younger respondents and narrows down as the age of respondents increases. However, these differences are small so the interaction appears to be slightly weak. 

### 4.4 Model 3: Logistic Regression Model Predicting Poverty Status Among Disabled Respondents

```{r, include=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
#Model 2 with binary VET_VAL
ProjectModel2 <- cleaned_pppub25 |>
  filter(ePRDISFLG == 1) |>
  dplyr::select(PERLIS, eA_SEX, binVET_VAL, DSAB_VAL, A_AGE, race_ethnicity, A_HGA_7cat, A_MARITL_3cat, A_LFSR_3cat, MIG_REG, ePEAFEVER) |>
  na.omit()

psych::describe(ProjectModel2)

lmodel2 <- glm(
  PERLIS ~ eA_SEX + binVET_VAL + DSAB_VAL + A_AGE + race_ethnicity + A_HGA_7cat + A_MARITL_3cat + A_LFSR_3cat + MIG_REG + ePEAFEVER,
  data = ProjectModel2,
  family = binomial()
)

summary(lmodel2)

```

```{r, include = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}

exp(cbind(OR = coef(lmodel2), confint(lmodel2)))

model_sum <- summary(lmodel2)

```

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.pos="H"}
table_g <- broom::tidy(lmodel2, conf.int = TRUE, exponentiate = FALSE) |>
  mutate(
    OR = exp(estimate),
    CI_lower = exp(conf.low),
    CI_upper = exp(conf.high),
    term = recode(
      term,
      "(Intercept)" = "Intercept",
      "eA_SEX1" = "Male",
      "binVET_VAL1" = "Veteran Payment",
      "ePEAFEVER" = "Veteran",
      "DSAB_VAL" = "Disability Income",
      "binVET_VALVeteran payment" = "Veteran payment",
      "ePRDISFLGDisability" = "Disability Status",
      "A_AGE" = "Age",
      "eA_SEXMale" = "Male",
      "A_MARITL_3catPreviously married" = "Previously married",
      "A_MARITL_3catNever married" = "Never married",
      "A_HGA_7catHigh School" = "High School",
      "A_HGA_7catHigh School Graduate" = "High School Graduate",
      "A_HGA_7catSome College" = "Some College",
      "A_HGA_7catAssociate's Degree" = "Associate's Degree",
      "A_HGA_7catBachelor's Degree" = "Bachelor's Degree",
      "A_HGA_7catAdvanced Degree" = "Advanced Degree",
      "race_ethnicityNon-Hispanic Black only" = "Non-Hispanic Black only",
      "race_ethnicityNon-Hispanic Asian only" = "Non-Hispanic Asian only",
      "race_ethnicityNon-Hispanic Indigenous only" = "Non-Hispanic Indigenous only",
      "race_ethnicityNon-Hispanic Multiracial/Other" = "Non-Hispanic Multiracial/Other",
      "race_ethnicityHispanic only" = "Hispanic only",
      "MIG_REGMidwest" = "Region: Midwest",
      "MIG_REGSouth" = "Region: South",
      "MIG_REGWest" = "Region: West",
      "MIG_REGAbroad" = "Region: Abroad",
      "A_LFSR_3catUnemployed" = "Unemployed",
      "A_LFSR_3catNot in labor force" = "Not in labor force"
    ),
    B = round(estimate, 3),
    SE = round(std.error, 3),
    z = round(statistic, 3),
    p = ifelse(p.value < .001, "< .001", sprintf("%.3f", p.value)),
    OR = round(OR, 3),
    `95% CI for OR` = paste0("[", round(CI_lower, 3), ", ", round(CI_upper, 3), "]")
  ) |>
  select(term, B, SE, z, p, OR, `95% CI for OR`) |>
  rename(Predictor = term)

kable(
  table_g,
  format = "latex",
  booktabs = TRUE,
  linesep = "",
  caption = "Government Assistance Predicting Poverty Status in Model 3"
) |>
  kable_styling(latex_options = c("HOLD_position", "scale_down")) |>
  footnote(general = "OR = odds ratio.")
```
In Table 6, the model's results indicate that government support, alongside demographic factors, significantly explain poverty risk among disabled individuals. After calculating odds ratios and 95% confidence intervals of the model's coefficients, analyzing each predictor variable becomes clearer. The model reveals how the effect of disability income (DSAB_VAL) is in fact statistically significant with a p-value of 2.42e-10, a beta value of -1.179e-04, a 95% CI of [0.99984, 0.99992], and an odds ratio of 0.99988. These results indicate that individuals receiving disability income will have 0.012% lower odds of poverty. Given that the confidence interval does not include 1, and that the p-value is less than 0.05, the effect of disability income is, again, statistically significant, with a consistent, protective effect. The model also reveals that the effect of veteran payments (binVET_VAL) is statistically significant with a p-value of 1.76e-13, a beta value of -1.470e+00, a 95% CI of [0.15, 0.33], and an odds ratio of 0.23. These results indicate that disabled individuals who receive veteran payments have roughly 77% lower odds of poverty. Given that the confidence interval does not include 1, and that the p-value is less than 0.05, the effect of veteran payments is, again, statistically significant, with a large protective effect. The variable veteran status (ePEAFEAVER) has a p-value of 0.2430, a beta value of -1.113e-01, a 95% CI of [0.74, 1.08], and an odds ratio of  0.89. Since the p-value is greater than 0.05, and the confidence interval includes 1, the effect of veteran identity status is not statistically significant. This can be attributed to the fact that poverty is better explained by financial support rather than veteran identity. Moving on to demographic effects, the model reveals that age is significantly significant with a p-value of < 2e-16, a beta value of -1.607e-02, a 95% CI of [0.981, 0.997], and an odds ratio of 0.98. The effect of age is small, but consistently protective with a ~1.6% decrease in the odds of poverty with every additional year of age among disabled individuals. 

When analyzing the effect of education (A_HGA_7cat), compared to base line, disabled individuals with some high school completion (p-value = 0.0163, Beta = -2.518e-01, 95% CI [0.63, 0.95], OR = 0.78) have 22% lower odds of poverty, disabled individuals who are high school graduates (p-value = 5.00e-08, Beta = -4.985e-01, 95% CI [0.51, 0.73], OR = 0.61) have 39% lower odds of poverty, disabled individuals with some college completion (p-value = 4.46e-08, Beta = -5.560e-01, 95% CI [0.47, 0.70], OR = 0.57) have 43% lower odds of poverty, disabled individuals with an associates degree (p-value = 3.17e-11, Beta = -7.693e-01, 95% CI [0.37, 0.58], OR = 0.46) have 54% lower odds of poverty, disabled individuals with a bachelor's degree (p-value = < 2e-16, Beta = -1.004e+00, 95% CI [0.29, 0.46], OR = 0.37) have 63% lower odds of poverty, lastly disabled individuals who have an advanced degree (p-value = < 2e-16, Beta = -1.306e+00, 95% CI [0.20, 0.36], OR = 0.27) have 73% lower odds of poverty. Since all p-values are less than 0.05, and all confidence intervals exclude 1, education is statistically significant with a strong protective effect. When analyzing the effect of race (race5), disabled individuals who identify as Black (p-value = < 2e-16, Beta =5.774e-01, 95% CI [1.58, 2.01], OR = 1.78) have 78% higher odds of poverty, disabled individuals who identify as Indigenous (p-value = 1.23e-05, Beta =5.997e-01, 95% CI [1.39, 2.38], OR = 1.82) have 82% higher odds of poverty. Since the p-values for Black and Indigenous disabled individuals are less than 0.05 and the confidence intervals exclude 1, identifying as Black or Indigenous has statistically significant effects on poverty. However, disabled individuals who identify as multiracial (p-value = 0.0668, Beta =2.403e-01, 95% CI [0.98, 1.64], OR = 1.27) and individuals who identify as Asian/Pacific Islander (p-value = 0.8194, Beta =2.806e-02, 95% CI [0.80, 1.30], OR = 1.03) do not have significant effects on poverty based on their race identities. 

When analyzing marital status (A_MARITL), disabled individuals who have been previously married (p-value = < 2e-16, Beta =1.149e+00, 95% CI [2.79, 3.57], OR = 3.16) and disabled individuals who have never been married (p-value = < 2e-16, Beta =9.903e-01, 95% CI [2.32, 3.12], OR = 2.69) both have  ~2.7–3.2× (169% to 216%) higher odds of poverty when compared to disabled married individuals. P-values are less than 0.05 and the confidence intervals exclude one indicating that the effects of marital status are statistically significant, having a strong and consistent effect on disabled individuals. When analyzing labor force status (A_LFSR_3cat), disabled individuals who are unemployed (p-value = 1.97e-15, Beta = 1.315e+00, 95% CI [2.68, 5.14], OR = 3.73) and disabled individuals who are not in the labor force (p-value = < 2e-16, Beta = 1.646e+00, 95% CI [4.39, 6.16], OR = 5.19) have over 5x (273% to 419%) higher odds of poverty when compared to disabled individuals who are part of the labor force. P-values are less than 0.05 and confidence intervals exclude 1 indicating that labor force status is statistically significant in its effects on poverty. Labor force status is one of the largest predictors in the model with the largest effects.

## 5. Discussion

### 5.1 Discussion of Model 1: Poverty Risk in the General Population

One of the major findings from Model 1 was that the additive version provided a better fit than the interaction, so the final model was estimated without the interaction between veteran payment status and disability status. The additive model had a slightly lower AIC than the interaction model, and the likelihood ratio test comparing the two models was not statistically significant. This suggested that veteran payment status did not meaningfully moderate the relationship between disability status and poverty in the general population. Instead, disability status and veteran payment status appeared to operate as independent predictors of poverty.

Model 1 also showed that veteran payment receipt was associated with substantially lower odds of poverty, while disability status remained associated with higher odds of poverty even after adjusting for demographic and socioeconomic controls. This pattern suggested that veteran-related financial support may have functioned as an important protective resource for some respondents, but that disability still carried an independent relationship with economic vulnerability. This finding was broadly consistent with prior scholarship showing that disability is often associated with poverty because of employment barriers, social exclusion, and unequal access to resources rather than solely because of individual health limitations [@Kopels01091995; @balazadeh2003disability; @heyman2024demographic]. At the same time, the negative association between veteran payment receipt and poverty echoed research suggesting that disability compensation and related federal support can function as an income floor that reduces material hardship, housing instability, and other forms of economic insecurity among veterans [@silver2026invisible; @edens2011association]. Rather than showing that compensation fully eliminated disadvantage, our results suggested a more limited buffering effect.

Model 1 also showed that poverty was strongly shaped by broader structural inequalities beyond disability and veteran payment status alone. Higher levels of education were consistently associated with lower odds of poverty, while unemployment and being out of the labor force were among the strongest predictors of elevated poverty risk. Marital status and race/ethnicity also mattered. Compared with married respondents, previously married and never married respondents had substantially higher odds of poverty, and several racial/ethnic minority groups had higher odds of poverty than Non-Hispanic White respondents. These findings aligned with broader research showing that poverty is tied to unequal educational opportunity, labor market exclusion, and racialized structural disadvantage [@heyman2024demographic; @Tamborini2019; @Sheehan2015]. In this sense, disability and veteran support did not operate in isolation. Instead, they appeared within a wider system of social and economic inequality that continued to shape poverty risk across the general population.

These findings also had practical and policy implications. The protective association between veteran payment receipt and poverty suggested that direct income support may have mattered in meaningful ways for reducing economic vulnerability. However, because disability remained positively associated with poverty, the results also suggested that compensation alone may not have been sufficient to offset the broader disadvantages associated with disability. This implied that anti-poverty efforts for disabled and veteran populations may have needed to go beyond benefit access alone and include stronger employment support, affordable health care, housing stability, and broader protections against uncertainty.

At the same time, Model 1 had several limitations. Since the analysis was cross-sectional, it did not allow us to make causal claims about whether veteran payment receipt directly reduced poverty risk. In addition, veteran payment status was modeled as a binary indicator rather than by payment amount, which likely obscured meaningful variation in the level of support respondents received. Future research should examine whether the amount, consistency, or type of veteran-related support changed the relationship between disability and poverty, and whether these patterns remained similar in longitudinal analyses that were better able to capture changes in poverty over time.

### 5.2 Discussion of Model 2:

Model 2 was an extension of model 1, which was whether or not veteran status and disability is associated with poverty status. The results from model 2 tells us whether or not veteran status moderates racial, age, disability status, and gender disparities in poverty status. 
Our hypothesis explores whether or not these demographic variables associated with veteran status affects poverty status. However, after performing a stepwise regression, we find that the model with only the age and veteran status interaction had the lowest AIC. We are limited to only finding only meaningful effects from the interaction between age and veteran status, but not the other interaction effects. The result was consistent with prior existing literature on Black adults, particularly in the younger age group, experienced higher rates of homelessness among both veterans and nonveterans[@Fargo2012]. Furthermore, the results also support the patterns that are consistent in the general population and the poverty population where veteran status was linked to a greater risk of homelessness[@Fargo2012]. Therefore, we partially accept that veteran status does moderate age and poverty status. However, we reject the hypothesis in which veteran status moderates race, sex, and disability because the interaction terms were not statistically significant. 

The limitations that comes with model 2 is the lack of significant demographic interactions after model testing. Further research is needed to expand on the effects of different demographics associated with veteran status on poverty status. Although the interaction between age and veteran status is statistically significant, the odds ratio is low which indicate that the effects of age differences are not large. Another limitation is that we don't have access to veteran's period of service which may lead to different poverty risks due to the difference in access to benefits, health status and economic conditions. Furthermore study should conduct the effects of poverty status based on the period that the veteran served. 


### 5.3 Discussion of Model 3:

Model 3 is focusing on the effect of government payments, both veteran payments and disability income, on poverty among disabled individuals. Our hypothesis explores how current government support and disability compensation helps protect disabled individuals from poverty. Reviewed literature presented a long history of disability being a key factor in systemic exclusion, particularly in the labor force, leading to disabled individuals experiencing major economic disadvantages [@Kopels01091995]. Findings from model 3 are consistent with literature on institutional barriers within social programs meant to protect disabled individuals from poverty [@heyman2024demographic]. Model 3 unveils how government support is in fact effective, but uneven. When filtering for only disabled individuals, veteran payments have a large meaningful reduction effect of poverty, where disabled income has a smaller marginal effect that is consistent with reducing the odds of poverty.  Thus, although government support is effective in its protection of disabled individuals from poverty, not all forms are equally effective. 

Model 3 also is consistent with literature on structural inequalities within the disabled population, indicating that protection against poverty is not solely reliant on monetary government support [@heyman2024demographic]. Disabled individuals who have received a higher education have more than half the odds of experiencing poverty than disabled individuals who have not received a higher education. Additionally, disabled individuals who are currently unmarried have higher odds of experiencing poverty when compared to married individuals, suggesting the positive economic benefits of marriage. When analyzing racial disparities among the disabled population, Black and Indigenous disabled individuals have more than double the odds of experiencing poverty when compared to other races. Lastly, the strongest predictor of poverty was labor force attachment. Disabled individuals who are not part of the labor force have five times the odds of experiencing poverty than disabled individuals who are part of the labor force, indicating the high vulnerability of disabled individuals who are not employed. 

Model 3’s findings have policy implications that can in turn lead to real practical, positive changes. The disproportionate effectiveness of veteran payments versus disability income is indicative of a larger issue of underfunding and the overlooking of flaws in systems. Given that veteran payments resulted in being highly protective, it should serve as a model for other government assistance programs for disabled individuals, ensuring that there is equity across all government programs in order to protect and support all disabled individuals. Furthermore, there should be a systemic focus on high-risk groups such as racial minorities, unemployed, and unmarried disabled individuals to ensure these vulnerable groups do not slip through the cracks. Moving away from policy surrounding income support, policies should be adapted to support the movement towards more integration of disabled individuals in the work force. Creating programs that directly encourage and facilitate employment for disabled individuals can have significant protective effects against poverty. 

It is important to note the limitations Model 3 has. For one, disability income and veteran payment is a fragmented measure for government support. There can be many supplemental resources that were not included in the data, preventing a full picture of the kind of aid and support each individual may receive. Additionally, there are many other factors that can lead to poverty such as cost of living, generational wealth, health severity, etc. that, again, were not included in this study, preventing a comprehensive analysis on poverty. Future research should take this into account, as well as include interaction terms to analyze the effects in more specific subgroups. Also interesting to include in future studies of this topic would be longitudinal data to more closely and accurately examine causality. 



\newpage

## References